# Practical Machine Learning

## Peer-graded Assignment: Prediction Assignment Writeup

The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.

Data download and required package loading
```{r}
library(dplyr)
library(caret)
library(ggplot2)
library(randomForest)

urlTrain <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
urlTest <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

download.file(urlTrain, destfile =  "./pml-training.csv")
download.file(urlTest, destfile =  "./pml-testing.csv")

training <- read.csv("pml-training.csv", na.strings=c("", "NA"))
testing <- read.csv("pml-testing.csv", na.strings=c("", "NA"))
unique(training$classe)
```

#### Data Exploratory Analysis
The str() and table() functions are used to understand the basic structure of the dataset. Due to the high number of columns (160), the result is subsetted:

```{r}
ncol(training)
```
```{r}
str(training[,1:10])
```
```{r}
str(training[,149:160])
```
```{r}
table(training$classe,training$user_name)
```
```{r}
ggplot(training, aes(classe)) + geom_bar(fill = "steelblue") + ggtitle("Counts per classe")
```

The plot shows that there is a relatively balanced distribution of observations among “classe” types.

#### Data pre-processing
The outcome “classe” must be converted into a factor variable. Additionally, there are many columns which do not provide any relevant information, because they either have plenty of NAs or because they are not actual predictors obtained from accelerator measurements. Those columns will be removed:
```{r}
training$classe <- as.factor(training$classe) # classe is converted into a factor variable.

trainingPrep <- training %>% select(8:160) # Non-predictors are removed.

trainingPrep <- trainingPrep %>% select_if(colSums(is.na(trainingPrep)) < 19000) # Only the columns with LESS than 19000 NAs are left (total nr. of obs. is 19622)

ncol(trainingPrep)
```

#### Create Data Partition
This dataset is further divided into train (75%) and test (25%) parts for cross-validation:
```{r}
inTrain = createDataPartition(trainingPrep$classe, p = 3/4)[[1]]
trainPart = trainingPrep[ inTrain,]
testPart = trainingPrep[-inTrain,]
```

#### Model training
A couple of models will be trained and tested with cross validation to find out which of them has the highest accuracy level. More precisely, a random forest model and an LDA model will be tested:

```{r}
set.seed(1234)
modfitrf <- randomForest(classe~., method = "class", data = trainPart)
predrf <- predict(modfitrf, newdata = testPart, type = "class")
confusionMatrix(predrf, testPart$classe)
```
```{r}
set.seed(1234)
modfitlda <- train(classe ~ ., method = "lda", data = trainPart)
predlda <- predict(modfitlda, newdata = testPart)
confusionMatrix(predlda, testPart$classe)
```

#### Model selection
The accuracy level of the random forest model (higher than 99%) is clearly higher than that of the LDA model (close to 70%). Therefore, the random forest model is selected.

####Cross validation and expected out of sample error
The out of sample error (calculated as 1 - Accuracy Level) is below 1%, therefore very low.

#### Prediction on 20 test cases
```{r}
predrf20 <- predict(modfitrf, newdata = testing, type = "class")
print(predrf20)
```
